---
title: "Tutorial: Descriptive Analysis for Network Graphs - hints"
author: "MSc in Statistics for Smart Data -- Introduction to graph analysis and modeling"
date: "Julien Chiquet, November the 6th, 2018"
fontsize: 11pt
lang: en
geometry: left=1.45in,top=1.35in,right=1.45in,bottom=1.35in
classoption: a4paper
linkcolor: red
urlcolor: blue
citecolor: green
output:
  pdf_document:
    number_sections: true
    citation_package: natbib
    includes:
      in_header: ../preamble.sty

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE)
```

# Packages requirements

```{r requirement}
library(tidygraph)
library(scholar)
library(igraph)
library(ggraph)
library(viridis)
```

# Network extraction

I found working with my co-authors' network was more fun: I use the **scholar** package to extract it.

```{r scholar network, cache = TRUE}
chiquet <- "FM2gRsYAAAAJ"
my_coauthors_network <- 
  scholar::get_coauthors(chiquet, n_coauthors = 10, n_deep = 1) %>% 
  graph_from_data_frame(directed = FALSE) %>% 
  igraph::simplify(remove.multiple = TRUE, remove.loops = TRUE)
```

Here is a fancy representation of this network with the **ggraph** package:
```{r ggraph scholar}
groups <- as.factor(as.numeric(membership(cluster_edge_betweenness(my_coauthors_network))))
my_coauthors_network %>% 
  tidygraph::as_tbl_graph() %>% activate(nodes) %>% 
  mutate(degree = centrality_degree(), 
         membership = groups,
         betweenness = centrality_betweenness()) %>% 
  ggraph() + scale_color_viridis() +
    geom_edge_link() + 
    geom_node_point(aes(size = degree, shape = membership, color = betweenness) ) + 
    geom_node_text(aes(label = name)) + 
    theme_void()
```

# Spectral Analysis

## Simple community network

Let us star by studying toy examples to better understand th spectral graph theory:

```{r community graph}
community <- igraph::sample_sbm(40, matrix(c(.25, .01, .01, .25), 2, 2), c(20, 20))
plot(community)
```


```{r laplacian}
L <- graph.laplacian(community, normalized = TRUE)
Matrix::image(L)
```

By, looking at space spawned by the first two eigen vector of $L$, we see that the clustering problem is outrageously easy (all point from the same community are superimposed) 

```{r eigen_vector}
U <- eigen(L)$vectors
Q <- 2
U <- U[,c((ncol(U)-Q+1):ncol(U))]
U <- U / rowSums(U^2)^(1/2)
plot(U[, 1], U[, 2])
```

## Spectral clustering of coauthorship network

Hereafter, an implementation of the spectral clustering algorithm studied during last session, 

```{r spectral_normalized}
SpectralClustering <- function (A,Q) {

  p <- ncol(A)
  if (Q > 1) {
    
    ## Normalized Laplacian
    L <- graph.laplacian(A, normalized = TRUE)    
    ## go into eigenspace
    U <- eigen(L)$vectors
    U <- U[,c((ncol(U)-Q+1):ncol(U))]
    U <- U / rowSums(U^2)^(1/2)
        
    ## Applying the K-means in the eigenspace
    cl <- kmeans(U, Q, nstart = 10, iter.max = 30)$cluster
  } else {
    cl <- as.factor(rep(1,ncol(A)))
  }
  cl
}
```

```{r clustering coaurthors}
cl_list <- lapply(2:7, function(Q)
  SpectralClustering(my_coauthors_network, Q)
)
A <- as_adj(my_coauthors_network)
lapply(cl_list, function(cl) 
  Matrix::image(
    A[order(cl), order(cl)],
    main = paste('number of cluster =',length(unique(cl)))
  )
)
```
